Species turnover reveals hidden effects of decreasing Nitrogen deposition in mountain hay meadows

### 2018-06-18 Manuscript.Rmd
Einleitung angepasst.

# Introduction {-}
Nitrogen deposition is the entry of reactive nitrogen compounds into water and soil via deposition through the air. Toghether with land-use and climate change, Nitrogen deposition is one of the major threats to biodiversity [@Sala2000; @Bobbink2010]. Since Nitrogen is an essential plant nutrient and many species-rich communities are adapted to conditions of low N availability [@Vitousek1997], the addition of Nitrogen strongly alter these communities [@Sala2000]. While there is strong evidence for the reduction in diversity of species-rich grasslands in Europe due to increased Nitrogen deposition [@Stevens2006; @Maskell2010; @Dupre2010], mountain grasslands have received less attention [@Humbert2016].

Since the late 1980s measures to reduce atmospheric pollution have successfully reduced emissions of nitrogen oxides, with affiliated decrease in Nitrogen deposition in many parts of Europe [@Torseth2012; @Fowler2007]. However, it is an open question whether and how fast the reduction in N deposition rates will lead to the recovery of extant plant communities (>>REF EINFÜGEN<<).

>> Kapitel zu konstante Communities und viel Turnover einfügen und dann erklären, dass wir eine Idee von Community assembly verwendet haben.

One useful approach to understanding biodiversity change is through estimates of biodiversity turnover reflecting both colonization and local extinction [@Hillebrand2018].

In Switzerland, grassland accounts for 70 per cent of the agricultural land. With extensive cultivation, permanent grassland has a very high biodiversity. This applies in particular to the meadows in the alpine region, where meadows with a high plant diversity are of great importance agronomically (Leiber et al. 2005), for the image of the products and for the conservation of biodiversity (Hohl, 2006). Observation data of vascular plants from field recordings shows negative correlations of Nitrogen deposition with species richness and species composition in mountain hay meadows [@Roth2013]. However, between 1990 and 2010 NOx emissions in Switzerland decreased by 46% and NH3 emissions by 14% (>>REF EINFÜGEN: Maas and Grennfelt, 2016<<), suggesting that mountain hay meadows should have recoverd.

However, mountain regions are threatened also by other global change drivers. In Switzerland, temperatures increased from 1959 to 2008 at all altitudes, with an average warming rate of 0.35&degC per decade, which is about 1.6 times the northern hemispheric warm- ing rate (Ceppi et al. 2010). In autumn and early winter, the warming rate depended on altitude with strongest trends at low altitudes, whereas in spring slightly stronger trends are found at altitudes close to the snow line. This climate warming is likely to directly affect or interact with Nitrogen deposition to drive plant community changes [@Humbert2016]. At the landscape scale, changes of plant communities were found that could be attributed to climate warming [@Roth2014]. Particularly, the shift of plant communities at mountain summits seems to be the restult of recent climate warming interacting with airborne nitrogen deposition [@Rumpf2018]. Furthermore, traditional management regimes are currently changing wich also has major impacts on grassland biodiversity in mountainous regions of Europe [@Niedrist2009; @Homburger2012]. In Switzerland  Management regimes of accessible mountainous areas are intensified, while poorly accessible mountainous areas are being abandoned [@Tasser2002; @Strebel2015].

Here, we used the data from the Swiss biodiversity monitoring [@Weber2004] and addressed the following questions:

>>1.  community measures , 2. Compare to turnover and 3. compare incicator values of species that newly colonized or disappeard (local extinction) from sites with random expectation.

@@

Nitrogen deposition at our plots decreased by `r mean(sites$NTOT_2015 - sites$NTOT_2000)` kg ha$^{-1}$ yr$^{-1}$ on average between 2000 and 2015. This is only about one tenth of the decrease in England, where Nitrogen deposition decreased by 24 kg ha$^{-1}$ yr$^{-1}$ from 1996 to 2011 [@Storkey2015]. This comparatively low change in Nitrogen deposition in Swiss mountain hay meadows might explain why we observed no change in average nutrient value of communities. Additionally, other anthropogenic pressures such as climate change might have overruled  Nitrogen depositin effects on community structure. Particularly, we found that the species disappearing from the sites tend to have below average indicator values for temperature, while species that newly colonize sites have above average indicator values for temperature (Table \@ref(tab:difffromrandomtab)). Thus, the effect of disappearing and the effect of colonizating species on the community mean for temperature is additive resulting in increasing average temperature values (Tab. \@ref(tab:communitytrendstab)). This is in contrast to how Nitrogen deposition is affecting disappearing and colonization of species: It seems that both species disappearing from the sites as well as species colonizing the sites tend to have below average indicator values for nutrients (Table \@ref(tab:difffromrandomtab)). Thus, the effect of disappearing and colonizating species on the average community value for nutrients partially cancel each other out. Furthermore, eutrophic species had rather high local survival across the entire deposition gradient, while oligotrophic species have much reduced local survival at high Nitrogen deposition. This suggests that it takes more time to replace eutrophic by oligotrophic species than replacing oligotrophic by eutrophic species. Taken toghether, these reasons might explain why the response of mountain hay medow communities responded stronger to to climate warming affects than to Nitrogen reduction, altough the reduction in Nitrogen resulted in above average colonization of oligotrophic species.

Observational studies along a gradient of Nitrogen deposition often infer how the spatial variation in species richness is related to Nitrogen deposition [@Stevens2010gradient; chapter 4 in @deVries2015]. They assume that the spatial variation in species richness (or other metrics of community structure) arose because over time some areas lost more species than others because they chronically experienced higher Nitrogen deposition. Although there is evidence supporting the use of such a 'space for time substitution' for detecting the effects of Nitrogen deposition on plant diversity [@Stevens2010], they can not replace studies that relate temporal patterns in species with Nitrogen deposition [@deSchrijver2011]. There are only a limited number of studies inferring temporal trends of plant species diversity related to varying amounts of N-deposition in extant communities [@Clark2008; @Storkey2015]. In an recent study, we used the mountain hay meadow data from a single survey only and estimated the critical load along the Nitrogen deposition gradient at which species richnes of oligotrophic species richness start to decrease with increasing Nitrogen deposition [@Roth2017]. Using this spatial variation in species richness and nitrogen deposition we estimated a critical load for mountain hay meadows of 13.1 kg ha$^{-1}$ yr$^{-1}$. In the current study we estimated the rate of Nitrogen deposition at which local survival probability or colonization probability was equal for oligotrophic and eutrophic species. This rate could also be interpreted as a critical load. With this aproach we obtained very similar results as the spatial approach. Our results are thus validating the space for time approach.

>>>>

# Introduction {-}
Nitrogen deposition is the entry of reactive nitrogen compounds into water and soil via deposition through the air. Since Nitrogen is an essential plant nutrient and many species-rich communities are adapted to conditions of low Nitrogen availability [@Vitousek1997], the addition of Nitrogen is likely to alter these communities. Indeed, Nitrogen deposition toghether with land-use and climate change is one of the major threats to biodiversity [@Sala2000; @Bobbink2010]. While there is strong evidence for the reduction in diversity of species-rich grasslands due to increased Nitrogen deposition [@Stevens2006; @Dupre2010; @Maskell2010; @Wesche2012], mountain grasslands have received less attention [@Humbert2016].

Since the late 1980s measures to reduce atmospheric pollution have successfully reduced emissions of nitrogen oxides, with affiliated decrease in Nitrogen deposition in many parts of Europe [@Torseth2012; @Fowler2007]. However, there are a number of factors that may prevent the recovery of plant communities that sufered from increased Nitrogen deposition. First, even if Nitrogen deposition is reduced large areas might still be above the critical load (>>REF<<). Second, there is a potential that communities reach an alternative stable state after decades of increased Nitrogen deposition and these species are unlikely to disappear again after the reduction of Nitrogen depositon [@Stevens2016]. Therefore, it is an open question whether and how fast the reduction in Nitrogen deposition rates will lead to the recovery of extant plant communities.

Recovry of extant plant communities from high Nitrogen depositon would imply that the state of communities measured at different time points is improving over time (i.e. improving biodiversity endpoints sensu @Rowe2017). Species richness is an easy to measure and communicate biodiversity entpoint and it is often negatively correlated with Nitrogen deposition [@Maskell2010; @Field2014; @Rowe2017]. Other metrics that are potentially more usful to reflect favourable changes can be derived from the traits of the species in a community. In Europe, environmental preference has often been expressed using incicator values assigned to each plant species [@Ellenberg1992]. Examples of such metrics would be the number of oligotrophic species or the average indicator value of the species in a community [@Roth2013; @Rowe2017]. However, the lack of a temporal trend in such biodiversity endpoints -- particularly in species richness -- does not necessarily mean that species composition remains unchanged. This is because immigration and extinction might be equally frequent and may cancel each other out [@Hillebrand2018]. Thus an useful approach to understanding biodiversity change is through estimates of biodiversity turnover reflecting both colonization and local extinction [@Hillebrand2018], especially if colonization and local extinction is compared to random expectation [@Chase2011].

In Switzerland, grassland accounts for 70 per cent of the agricultural land. With extensive cultivation, permanent grassland has a very high biodiversity. This applies in particular to the meadows in the alpine region, where meadows with a high plant diversity are of great importance agronomically (Leiber et al. 2005), for the image of the products and for the conservation of biodiversity (Hohl, 2006). Observation data of vascular plants from field recordings shows negative correlations of Nitrogen deposition with species richness and species composition in mountain hay meadows [@Roth2013]. Between 1990 and 2010 NOx emissions in Switzerland decreased by 46% and NH3 emissions by 14% (>>REF EINFÜGEN: Maas and Grennfelt, 2016<<), suggesting that Nitrogen deposition has decreased also in mountain hay meadows should and we expected plant communities to respond accordingly.

However, Swiss mountain ecosystems are threatened also by other global change drivers. In Switzerland, temperatures increased from 1959 to 2008 at all altitudes, with an average warming rate of 0.35$^{\circ}$C per decade, which is about 1.6 times the northern hemispheric warm- ing rate [@Ceppi2012]. This climate warming is likely to directly affect or interact with Nitrogen deposition to drive plant community changes [@Humbert2016]. Indeed, in an earlier study we found that at a landscape scale plant communities responded to climate warming even within a comparably short time period [@Roth2014] and particularly the shift of plant communities at mountain summits are the restult of recent climate warming interacting with airborne nitrogen deposition [@Rumpf2018]. Furthermore, traditional management regimes are currently changing wich also has major impacts on plant communities in mountainous regions of Europe [@Niedrist2009; @Homburger2012]. In Switzerland  Management regimes of accessible mountainous areas are intensified, while poorly accessible mountainous areas are being abandoned [@Tasser2002; @Strebel2015].

Here, we used the data from the Swiss biodiversity monitoring [@Weber2004] and addressed the following questions: (1) Did biodiversity endpoints that are likely to reflect temperature, precipitation, nitrogen deposition or land-use intensity change over the last 15 years, (2) Was species turnover correlated with the avearage temperature, precipitation, Nitrogen deposition or stepness (we expect step areas to be less intensively managed), and (3) did species that newly colonized or disappeard (local extinction) from sites different from random expectation according to their incicator values for tempearture, humidity, nutrients or light.

@@

Nitrogen deposition at our plots decreased by `r mean(sites$NTOT_2015 - sites$NTOT_2000)` kg ha$^{-1}$ yr$^{-1}$ on average between 2000 and 2015. This is only about one tenth of the decrease in England, where Nitrogen deposition decreased by 24 kg ha$^{-1}$ yr$^{-1}$ from 1996 to 2011 [@Storkey2015]. This comparatively low change in Nitrogen deposition in Swiss mountain hay meadows might explain why we observed no change in average nutrient value of communities. Additionally, other anthropogenic pressures such as climate change might have overruled  Nitrogen depositin effects on community structure. Particularly, we found that the species disappearing from the sites tend to have below average indicator values for temperature, while species that newly colonize sites have above average indicator values for temperature (Table \@ref(tab:difffromrandomtab)). Thus, the effect of disappearing and the effect of colonizating species on the community mean for temperature is additive resulting in increasing average temperature values (Tab. \@ref(tab:communitytrendstab)). This is in contrast to how Nitrogen deposition is affecting disappearing and colonization of species: It seems that both species disappearing from the sites as well as species colonizing the sites tend to have below average indicator values for nutrients (Table \@ref(tab:difffromrandomtab)). Thus, the effect of disappearing and colonizating species on the average community value for nutrients partially cancel each other out. Furthermore, eutrophic species had rather high local survival across the entire deposition gradient, while oligotrophic species have much reduced local survival at high Nitrogen deposition. This suggests that mountain hay meadow communities can reach alternative stable states with eutrophic species that are unlikely to disappear again even if Nitrogen depositon is decreasing again [@Stevens2016]. Taken toghether, these reasons might explain why the response of mountain hay medow communities responded stronger to to climate warming affects than to Nitrogen reduction, altough the reduction in Nitrogen resulted in above average colonization of oligotrophic species.

Observational studies along a gradient of Nitrogen deposition often infer how the spatial variation in species richness is related to Nitrogen deposition [@Stevens2010gradient; chapter 4 in @deVries2015]. They assume that the spatial variation in species richness (or other metrics of community structure) arose because over time some areas lost more species than others because they chronically experienced higher Nitrogen deposition. Although there is evidence supporting the use of such a 'space for time substitution' for detecting the effects of Nitrogen deposition on plant diversity [@Stevens2010], they can not replace studies that relate temporal patterns in species with Nitrogen deposition [@deSchrijver2011]. There are only a limited number of studies inferring temporal trends of plant species diversity related to varying amounts of N-deposition in extant communities [@Clark2008; @Storkey2015; @Stevens2016]. In an recent study, we used the mountain hay meadow data from a single survey only and estimated the critical load along the Nitrogen deposition gradient at which species richnes of oligotrophic species richness start to decrease with increasing Nitrogen deposition [@Roth2017]. Using this spatial variation in species richness and nitrogen deposition we estimated a critical load for mountain hay meadows of 13.1 kg ha$^{-1}$ yr$^{-1}$. In the current study we estimated the rate of Nitrogen deposition at which local survival probability or colonization probability was equal for oligotrophic and eutrophic species. This rate could also be interpreted as a critical load. With this aproach we obtained very similar results as the spatial approach. Our results are thus validating the space for time approach.
